Direct Entropy Measurement in a Mesoscopic Quantum System

### 2018-02-06 paper/draft.tex
some figure updates and a few revisions

\documentclass[twocolumn,showpacs,preprintnumbers,amsmath,amssymb,pra,aps,superscriptaddress]{revtex4-1}

@@

\title{Direct Entropy Measurement in a Mesoscopic Quantum System}
\author{Nikolaus Hartman}
\email{nik.hartman@gmail.com}
\thanks{The raw data and Python code used for the figures and analysis can be found at \url{https://github.com/nikhartman/spin_entropy}}
\affiliation{University of British Columbia, Vanouver, BC, Canada}
\author{Saeed Fallahi}
\affiliation{Purdue University, Lafayette, IN, USA}
\author{Geoffrey C. Gardner}
\affiliation{Purdue University, Lafayette, IN, USA}
\author{Christian Olsen}
\affiliation{University of British Columbia, Vanouver, BC, Canada}
\author{Silvia Folk}
\affiliation{University of British Columbia, Vanouver, BC, Canada}
\author{Mohammad Samani}
\altaffiliation{The Hospital for Sick Children, Toronto, ON, Canada}
\altaffiliation{Fields Institute for Research in Mathematical Sciences, Toronto, ON, Canada}
\affiliation{University of British Columbia, Vanouver, BC, Canada}
\author{Michael Manfra}
\affiliation{Purdue University, Lafayette, IN, USA}
\author{Joshua Folk}
\affiliation{University of British Columbia, Vanouver, BC, Canada}
\date{\today}

@@

Measuring the entropy of an electronic state is a powerful tool for identifying its underlying microscopic character.  Such measurements are typically based on bulk properties, such as heat capacity, that are straightforward to observe in macroscopic samples but exceedingly difficult to access in mesoscopic systems that may consist of just a few electrons. Taking advantage of a well-known Maxwell relation, we realize a protocol for entropy-to-charge conversation in a gate-defined GaAs quantum dot that enables an entropy measurement of the first three quantum states in to the dot. The entropy of a single spin ($k_B \ln{2}$) is measured within 8\% accuracy, as is the entropy arising at the magnetic field-driven singlet-triplet crossing for two electrons.

@@

%%% figure 1 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_1_no-annotation.pdf}
        \caption{\label{fig:fig1}(a) Scanning electron micrograph of a device similar to the one measured, showing electrostatic gates (blue) used to define a circuit in the 2D electron gas. Grey gates are unused (grounded). Squares show the locations of ohmic contacts to the 2DEG. An electron reservoir is formed between the parallel gates running vertically down the image, with a 200nm diameter quantum dot coupled to the right side. The channel can be heated by driving current through the quantum point contact on the left. Electrons tunneling between the reservoir and dot are measured with a capacitively-coupled charge sensor. (b)  Charge sensor signal showing a single transition from $N \rightarrow N+1$ electrons at two temperatures ($T_{Red} > T_{Blue}$). (c) Change in the charge sensor signal with respect to temperature. The asymmetry in peak height arises from a shift in $V_0$ that would arise from a change in entropy between the $N$ and $N+1$ electron states. In the case of $dS=0$, $dg$ is anti-symmetric about $V_0$.}
\end{figure}

@@

%%% figure 2 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_2.pdf}
        \caption{\label{fig:fig2}(a) Charge sensor data for $N=0 \rightarrow 1$ at 100 and 160mK. Temperature is set by DC current through the QPC heater. (b) Lockin measurement of $dg_{sens}$ with $dT \sim 60mK$. Fits to $dg$ are shown with $\delta$ as a free parameter (solid) and $\delta=0$ (dashed) (c) Transition width, $\Theta$, as a function of mixing chamber temperature. The highlighted, linear region is where the tunnel rates are determined only by temperature broadening. The lever arm $\alpha$ is calculated by fitting a straight line to this region. (d) $\Theta$ as a function of DC current through the QPC heater. Linear fit is used to convert between $I_{heat}$ and $dT$, with $R_{QPC} = $ \SI{20}{\kilo\ohm} (e) Change in entropy as a function of AC current through the QPC heater. Top axis shows corresponding $dT$. Right axis shows shift in $V_0$ per unit temperature.}
\end{figure}

@@

%%% figure 3 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_3.pdf}
        \caption{\label{fig:fig3}(a) Transport data through the quantum dot showing the $N=0 \rightarrow 1$ transition. The first excited state, near $V_{SD} = $ \SI{1}{\milli\electronvolt}, is well-above all other energy scales in the measurement. Dashed line at $V_{SD}$ = \SI{700}{\micro\electronvolt} shows where data in (b) are taken. (b) Fixed bias transport data showing fits to the Zeeman splitting of the ground state energy from which $g = -0.44$ is extracted. (c) Change in entropy, determined from $dg$ fits at varying parallel magnetic field. The $N=0 \rightarrow 1$ and $2 \rightarrow 3$ data are overlaid here to show the similar behavior of these states at low field. (d) and (e) show characteristic $dg$ traces from which the data in (c) were extracted. These data points are show as large markers in (c)}
\end{figure}

@@

%%% figure 4 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_4.pdf}
        \caption{\label{fig:fig4}(a) (a) Transport data through the quantum dot showing the $N=1 \rightarrow 2$ transition. Singlet-triplet splitting is roughly \SI{300}{\micro\electronvolt} and depends on applied field. Dashed line at $V_{SD}$ = \SI{1250}{\micro\electronvolt} shows where data in (b) are taken. (b) Fixed bias transport data in parallel field. The triplet level is split into $T_+$, $T_0$, and $T_{-}$ (not visible). At \SI[input-protect-tokens]{\sim 9}{\tesla} $T_+$ becomes degenerate with $S$. $g$ is determined using $T_0$ and $T_+$ fits. (c) Change in entropy, extracted from $dg$ fits at varying parallel field. As the field goes from 0 to 10T, entropy of the 2-electron state goes from $0  \rightarrow \ln{2}$, while the entropy of the 1-electron state goes from $\ln{2} \rightarrow 0$. Fit to this behavior (dashed) is described in the text (d), (e), and (f) show characteristic $dg$ traces from which the data in (c) were extracted. These data points are show as large markers in (c)}
\end{figure}

@@

The field-dependent entropy measurement for the 1-2 transition can again be fit using Eq.~\ref{eqn:gibbs}, with probabilities as before for the 1-electron states and $p_{S/T}(B_\parallel, T) = (1+ e^{\mp \frac{g\mu_B B_\parallel - \Delta_{ST}}{k_B T}})^{-1}$ for the 2-electron singlet-triplet states, where $\Delta_{ST}$ is the singlet-triplet splitting at zero field. Fitting the data in Fig.~\ref{fig:fig4}c we find $g(B{=}0T)=-0.46$, $g(B{=}9T)=-0.31$, $\Delta_{ST}(B{=}9T)$ = \SI{240}{\micro\electronvolt}, $dS(B{=}0T) = 1.04\ln{2}$, and $dS(B{=}9T) = 1.01\ln{2}$. Each of these values is consistent with data in Figs.~\ref{fig:fig4}a and b as well as previous GaAs dot measurements [cite why g should change in high field]. This fit confirms that the $dS$ signal can be related directly to a tunable degeneracy in the system.

@@

\noindent \textbf{Methods} The device was built on a AlGaAs/GaAs heterostructure, hosting a 2D electron gas with density and mobility at \SI{300}{\milli\kelvin} (determined on a separate chip) of \SI{2.42e11}{\per\square\centi\metre} and \SI[per-mode=symbol]{2.56e6}{\square\centi\metre\per\volt\per\second}.   Mesas and NiAuGe ohmic contacts to the 2DEG were defined by standard photolithography techniques, then \SI{10}{\nano\metre} of $\mathrm{HfO_2}$ was deposited via atomic layer deposition to improve the gating stability to the device. Electron beam lithography followed by evaporation of \SI{20}{\nano\metre} of Ti/Au defined the fine gate structures, including the \SI{200}{\nano\metre} diameter quantum dots. The measurement was carried out in an dilution refrigerator with a base temperature of \SI{14}{\milli\kelvin}.

The conductance of the charge sensor was measured at a DC voltage bias of XXX uV, with the current read out directly or fed into a lockin amplifier.  The temperature of the reservoir could be raised above $T_MC$ (to $T_MC + dT$) using a current biased through the left (heater) QPC.  DC heating currents were used to calibrate heating efficiency in this mode, while AC currents at $f_{Heat} =$ \SI{48.7}{\hertz} yielded oscillations in $T_{res}$ at \SI{97.4}{\hertz}.  Locking the current measurement through the sensor QPC into the second harmonic of $f_{Heat}$ thus provided a measurement of the variation of $g_{sens}$ with $dT$. A calibration for $dT$, using $\alpha$ as measured in Fig.~\ref{fig:fig2}b, can be seen in Fig.~\ref{fig:fig2}c.

>>>>

%%%% outstanding mysteries %%%%%
% error bar on 0-1 peak height (mention in text)
% can I cite someone as to why \Delta_ST changes with field
% dS or \partial S?
% why does \partial S/ \partial N

\documentclass[preprint,showpacs,preprintnumbers,amsmath,amssymb,pra,aps,superscriptaddress]{revtex4-1}

@@

\title{Direct Entropy Measurement in a Mesoscopic Quantum System}
\author{Nikolaus Hartman}
\email{nik.hartman@gmail.com}
\thanks{The raw data and Python code used for the figures and analysis can be found at \url{https://github.com/nikhartman/spin_entropy}}
\affiliation{University of British Columbia, Vancouver, BC, Canada}
\author{Saeed Fallahi}
\affiliation{Purdue University, Lafayette, IN, USA}
\author{Geoffrey C. Gardner}
\affiliation{Purdue University, Lafayette, IN, USA}
\author{Christian Olsen}
\affiliation{University of British Columbia, Vancouver, BC, Canada}
\author{Silvia Folk}
\affiliation{University of British Columbia, Vancouver, BC, Canada}
\author{Mohammad Samani}
\altaffiliation{The Hospital for Sick Children, Toronto, ON, Canada}
\altaffiliation{Fields Institute for Research in Mathematical Sciences, Toronto, ON, Canada}
\affiliation{University of British Columbia, Vancouver, BC, Canada}
\author{Michael Manfra}
\affiliation{Purdue University, Lafayette, IN, USA}
\author{Joshua Folk}
\affiliation{University of British Columbia, Vancouver, BC, Canada}
\date{\today}

@@

%Measuring the entropy of an electronic state is a powerful tool for identifying its underlying microscopic character.  Such measurements are typically based on bulk properties, such as heat capacity, that are straightforward to observe in macroscopic samples but exceedingly difficult to access in mesoscopic systems that may consist of just a few electrons. Taking advantage of a well-known Maxwell relation, we realize a protocol for entropy-to-charge conversation in a gate-defined GaAs quantum dot that enables an entropy measurement of the first three quantum states in to the dot. The entropy of a single spin ($k_B \ln{2}$) is measured within 8\% accuracy, as is the entropy arising at the magnetic field-driven singlet-triplet crossing for two electrons.

@@

%%% figure 1 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_1_no-annotation.pdf}
        \caption{\label{fig:fig1}(a) Scanning electron micrograph of a device similar to the one measured, showing electrostatic gates (blue) used to define a circuit in the 2D electron gas. Grey gates are unused (grounded). Squares show the locations of ohmic contacts to the 2DEG. An electron reservoir is formed between the parallel gates running vertically down the image, with a 200nm diameter quantum dot coupled to the right side. The channel can be heated by driving current through the quantum point contact on the left. Electrons tunneling between the reservoir and dot are measured with a capacitively-coupled charge sensor. (b)  Charge sensor signal showing a single transition from $N \rightarrow N+1$ electrons at two temperatures ($T_{Red} > T_{Blue}$). (c) Change in the charge sensor signal with respect to temperature. The asymmetry in peak height arises from a shift in $V_0$ that would arise from a change in entropy between the $N$ and $N+1$ electron states. In the case of $dS=0$, $dg_{sens}$ is anti-symmetric about $V_0$.}
\end{figure}

@@

%%% figure 2 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_2.pdf}
        \caption{\label{fig:fig2}(a) Charge sensor data for $N=0 \rightarrow 1$ at 100 and 160mK. Temperature is set by DC current through the QPC heater. (b) Lockin measurement of $dg_{sens}$ with $dT \sim 60mK$. Fits to $dg_{sens}$ are shown with $\delta$ as a free parameter (solid) and $\delta=0$ (dashed) (c) Transition width, $\Theta$, as a function of mixing chamber temperature. The highlighted, linear region is where the tunnel rates are determined only by temperature broadening. The lever arm $\alpha$ is calculated by fitting a straight line to this region. (d) $\Theta$ as a function of DC current through the QPC heater. Linear fit is used to convert between $I_{heat}$ and $dT$, with $R_{QPC} = $ \SI{20}{\kilo\ohm} (e) Change in entropy as a function of AC current through the QPC heater. Top axis shows corresponding $dT$. Right axis shows shift in $V_0$ per unit temperature.}
\end{figure}

@@

%%% figure 3 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_3.pdf}
        \caption{\label{fig:fig3}(a) Transport data through the quantum dot showing the $N=0 \rightarrow 1$ transition. The first excited state, near $V_{SD} = $ \SI{1}{\milli\electronvolt}, is well-above all other energy scales in the measurement. Dashed line at $V_{SD}$ = \SI{700}{\micro\electronvolt} shows where data in (b) are taken. (b) Fixed bias transport data showing fits to the Zeeman splitting of the ground state energy from which $g = -0.44$ is extracted. (c) Change in entropy, determined from $dg_{sens}$ fits at varying parallel magnetic field. The $N=0 \rightarrow 1$ and $2 \rightarrow 3$ data are overlaid here to show the similar behavior of these states at low field. (d) and (e) show characteristic $dg_{sens}$ traces from which the data in (c) were extracted. These data points are show as large markers in (c)}
\end{figure}

@@

%%% figure 4 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_4.pdf}
        \caption{\label{fig:fig4}(a) (a) Transport data through the quantum dot showing the $N=1 \rightarrow 2$ transition. Singlet-triplet splitting is roughly \SI{300}{\micro\electronvolt} and depends on applied field. Dashed line at $V_{SD}$ = \SI{1250}{\micro\electronvolt} shows where data in (b) are taken. (b) Fixed bias transport data in parallel field. The triplet level is split into $T_+$, $T_0$, and $T_{-}$ (not visible). At \SI[input-protect-tokens]{\sim 9}{\tesla} $T_+$ becomes degenerate with $S$. $g$ is determined using $T_0$ and $T_+$ fits. (c) Change in entropy, extracted from $dg_{sens}$ fits at varying parallel field. As the field goes from 0 to 10T, entropy of the 2-electron state goes from $0  \rightarrow \ln{2}$, while the entropy of the 1-electron state goes from $\ln{2} \rightarrow 0$. Fit to this behavior (dashed) is described in the text (d), (e), and (f) show characteristic $dg_{sens}$ traces from which the data in (c) were extracted. These data points are show as large markers in (c)}
\end{figure}

@@

The field-dependent entropy measurement for the 1-2 transition can again be fit using Eq.~\ref{eqn:gibbs}, with probabilities as before for the 1-electron states and $p_{S/T}(B_\parallel, T) = (1+ e^{\mp \frac{g\mu_B B_\parallel - \Delta_{ST}}{k_B T}})^{-1}$ for the 2-electron singlet-triplet states, where $\Delta_{ST}$ is the singlet-triplet splitting at zero field. The fit allows for an offset from $dS=0$, away from the degenerate points, to compensate for non-linearities in the charge sensor. From the fit we find the entropies at the two-fold degenerate points, $B=0$ and \SI{9}{\tesla}, are within 5\% of the expected values, $\mp k_B \ln{2}$. The extracted value of $g = 0.394$ from the low-field peak agrees well with the fit in Fig.~\ref{fig:fig3}a. At the high-field singlet-triplet degeneracy, $g = 0.586$ in the fit. This unexpected g-factor is explained by a shifting of the $T_{0}$ with magnetic field, as seen in Fig.~\ref{fig:fig4}f and previous work. The high quality of this Gibbs entropy fit confirms $dS$ can be related directly to a tunable degeneracy in the system.

@@

\paragraph*{Methods} The device was built on a AlGaAs/GaAs heterostructure, hosting a 2D electron gas with density and mobility at \SI{300}{\milli\kelvin} (determined on a separate chip) of \SI{2.42e11}{\per\square\centi\metre} and \SI[per-mode=symbol]{2.56e6}{\square\centi\metre\per\volt\per\second}.   Mesas and NiAuGe ohmic contacts to the 2DEG were defined by standard photolithography techniques, then \SI{10}{\nano\metre} of $\mathrm{HfO_2}$ was deposited via atomic layer deposition to improve the gating stability to the device. Electron beam lithography followed by evaporation of \SI{20}{\nano\metre} of Ti/Au defined the fine gate structures, including the \SI{200}{\nano\metre} diameter quantum dots. The measurement was carried out in an dilution refrigerator with a base temperature of \SI{14}{\milli\kelvin}.

The conductance of the charge sensor was measured at a DC voltage bias of XXX uV, with the current read out directly or fed into a lockin amplifier.  The temperature of the reservoir could be raised above $T_MC$ (to $T_MC + dT$) using a current biased through the left (heater) QPC. A calibration for $dT$ in this mode, using DC current bias $\alpha$ as measured in Fig.~\ref{fig:fig2}b, can be seen in Fig.~\ref{fig:fig2}c. Applying AC current at $f_{Heat} =$ \SI{48.7}{\hertz} yields oscillations in $T_{res}$ at $2f_{heat}$; to leading order $dT_{res} \sim P \sim \cos(2*(2 \pi f_{heat})t)$. Variations in $g_{sens}$ with $dT$ are captured with a lockin amplifier measuring $g_{sens}$ oscillations at at the second harmonic of $f_{heat}$.
